\mytitle

### 2020-04-06 msc_thesis.tex
re-structure msc thesis code

\documentclass[11pt,a4paper]{article}
% Font
\usepackage{helvet}
\renewcommand{\familydefault}{\sfdefault}
% Encoding and language
\usepackage[utf8]{inputenc}
\usepackage[english]{babel}
% Colors
\usepackage[usenames,dvipsnames]{color}
\usepackage{colortbl}
% Paper size and margins
%\usepackage{rotating}
\usepackage{pdflscape}
%\usepackage[a4paper,showframe]{geometry}
\usepackage{vmargin}
\setmarginsrb{2.5cm}{2cm}{2.5cm}{2cm}{0cm}{0cm}{0cm}{1.5cm}
\usepackage{pdflscape}
\usepackage{afterpage}
% Indenting and line spacing
\linespread{1.5}
%\usepackage{setspace}
\usepackage{indentfirst}
%\setlength{\parindent}{1cm}
\setlength{\parskip}{0.5cm}
% Symbols
\usepackage{amsmath}
\usepackage{amsfonts}
\usepackage{amssymb}
% Equations
\usepackage{eulervm} % upright font in eqns (Palatino)
\usepackage{cool}
\usepackage{mathtools} % for arrows

\newcounter{matriz}
\newenvironment{matriz}{\refstepcounter{matriz}\equation}{\tag{\thematriz}\endequation}
\renewcommand{\theequation}{R\arabic{equation}}

% References
\usepackage{natbib}
\bibliographystyle{apalike}
% Hyperlinks and labels
%\usepackage{showkeys}
\usepackage[linktocpage=true,plainpages=false,pdfpagelabels=false]{hyperref}
\definecolor{citecolor}{rgb}{0.2,0.7,0.9}
\definecolor{urlcolor}{rgb}{0,0,1}
\hypersetup{
    colorlinks, linkcolor={Blue},
    citecolor={Blue}, urlcolor={Black}
}
% Figures
\usepackage{subcaption}
\usepackage{alphalph}
\usepackage{tikz}
\usetikzlibrary{shapes,arrows}
\usetikzlibrary{positioning}
\usetikzlibrary{calc}
\usepackage{epstopdf}
\usepackage{wrapfig}
% Tables
\usepackage{multirow}
\usepackage{tabularx}
\usepackage{array}
\usepackage{ragged2e}
\newcolumntype{P}[1]{>{\RaggedRight\hspace{0pt}}p{#1}}
% Tikz styles
\tikzstyle{reg} = [rounded rectangle, text width=2cm, fill=black!5, align=center, anchor=west,font=\sffamily\Large\bfseries, minimum height=1cm, inner sep=0]
\tikzstyle{rad} = [rounded rectangle, draw=none, text width=2cm, fill=red!10, align=center, anchor=west,font=\sffamily\Large\bfseries, minimum height=1cm, inner sep=0]
\tikzstyle{long} = [rounded rectangle, draw=none, text width=3.5cm, fill=black!5, align=center, anchor=west,font=\sffamily\Large\bfseries, minimum height=1cm, inner sep=0]   
\tikzstyle{add} = [scale=1.5, draw=none,fill=none,align=center]
\tikzstyle{emp} = [draw=none,fill=none]
\tikzstyle{ar} = [->,draw=black!70,line width=2]
\tikzstyle{ln} = [-,draw=black!70,line width=2]
% Title etc
\newcommand{\mytitle}{Investigation of the relationship between tropospheric ozone production efficiency and carbon bond emissions}
\newcommand{\authorname}{Maria Zamyatina}
\newcommand{\authornumber}{Student No: 100106685}
\newcommand{\degname}{Master of Science}
\newcommand{\univname}{University of East Anglia}
\newcommand{\univaddr}{University Plain \\ Norwich NR47TJ UK}
\newcommand{\depname}{School of Environmental Sciences}
\newcommand{\mydate}{\the\year}

\newcommand{\picspath}{D:/UEA/ENV-MB4X_Dissertation/ENV-MB4X_Dissertation_thesis/ENV-MB4X_Dissertation_thesis_latex/pics/}
%% Remove author and date from the maketitle command
%\makeatletter
%\renewcommand{\@maketitle}{
%\newpage
% \null
% \vskip 2em%
% \begin{center}%
%  {\LARGE \@title \par}%
% \end{center}%
% \par} \makeatother

% PDF meta-data
\hypersetup{pdftitle={\mytitle}}
\hypersetup{pdfauthor=\authorname}

\title{\mytitle}
% =========================================================================
\begin{document}

\input{diss_titlepage}
%\onehalfspacing

@@

\newpage
\tableofcontents
\newpage

\section{Introduction} \label{sec:intro}
The Earth’s atmosphere is the recipient of a vast range of chemical compounds emitted by natural and anthropogenic sources. The oxidation of these compounds is the main process taking place in the atmosphere that keeps their concentrations at levels that do not distort the chemical balance of the atmosphere \citep{Prinn2003}. In other words, oxidation is a removal 'cleansing' mechanism for these compounds (various pollutants and greenhouse gases), and its rate is often referred to as the oxidizing capacity of the atmosphere. This capacity depends mainly on two factors: climate and the global burden of the principal oxidants such as ozone ($O_3$), the hydroxyl radical ($OH$) and hydrogen peroxide ($H_2O_2$) \citep{Prinn2003,Thompson1992}.
\subsection{Tropospheric ozone}\label{sec:into_O3}
Tropospheric ozone ($O_3$) plays an important role in the chemical and thermal balance of the atmosphere. Although it accounts only for about 10\% of the total ozone, it is a primary precursor to the formation of hydroxyl radicals ($OH$) which in association with hydrogen peroxide ($H_2O_2$) determines the 'cleansing' capacity of the atmosphere \citep{Prinn2003,Tarasick2008,Thompson1992}. On the other hand, being a source of main pollutant removal agents, ozone is an air pollutant itself. Long exposure to its high concentrations causes respiratory problems in humans and damage to sensitive plant species \citep{Fowler2008}. Last but not least characteristic of tropospheric ozone is that it is an important greenhouse gas with a current estimated radiative forcing of  $0.40\pm 0.20~Wm^{–2}$ which is approximately one fifth of the $CO_2$ radiative forcing \citep{Hartmann2013,Myhre2013}. Due to these reasons a considerable interest in quantifying tropospheric ozone concentrations, budget and trends exists for more than a 100 years \citep{Becker2004}.

Unlike many other air pollutants, tropospheric ozone is not directly emitted, but produced following the oxidation of carbon monoxide ($CO$), methane ($CH_4$), and nonmethane volatile compounds (nmVOCs) in the presence of nitrogen oxides ($NO_x$) \citep{Crutzen1973,Myhre2013}. The emissions of these precursors arise from both natural and anthropogenic sources which have intensified since pre-industrial times \citep{Parrish2014,Volz1988} (Figure \ref{fig:O3observations}). Presumably it has led to roughly a doubling in background tropospheric ozone concentration \citep{Guicherit2000,Hartmann2013,Tarasick2008,Vingarzan2004}, although contribution from another major source, transport from the stratosphere, may also took place \citep{Fowler2008}. Climate is also believed to be a key driver of ozone production and destruction since the chemical reactions affecting ozone concentration are strongly dependent on such climatic factors as temperature, rainfall and humidity. Therefore in a changing climate tropospheric ozone is not anymore a local air quality issue, but is a global pollution problem \citep{Fowler2008}.

\begin{figure}[h]
\centering
\includegraphics[width=\linewidth]{{./pics/Isaksen2009_O3}.png}
\caption{Observed surface ozone at different Northern Hemispheric surface stations \citep{Isaksen2009}.}
\label{fig:O3observations}
\end{figure}

According to the Intergovernmental Panel on Climate Change (IPCC) Fifth Assessment report (AR5) in the present conditions (around year 2000) the global mean tropospheric ozone budget is approximately 331 Tg \citep{Myhre2013}. However, despite general agreement on how the drivers impact ozone concentrations at global and regional scales, there is considerable inconsistency in its budget between different modelling and observational studies \citep{Stevenson2006,Wild2007,Young2012}. Figure \ref{fig:O3budget} demonstrates that the range of global tropospheric ozone budget estimates is approximately 60 Tg which is 18\% of the mean. In terms of ozone production the variation is even higher and accounts for 28\% of the corresponding value. In that regard, two questions arise: why does the tropospheric ozone budget differ between models and observations? And secondly, why does it vary to this extent, especially in terms of production? The only way to answer these questions is to perform a rigorous investigation of the factors that drive tropospheric ozone and attribute those differences to respective factors \citep{Young2012}.

\begin{figure}[t] % O3 budget
\centering
\includegraphics[width=0.8\linewidth]{o3_studies.eps}
\caption{Summary of tropospheric ozone global budget model and observation estimates for present (about 2000) conditions. All uncertainties quoted as 1 standard deviation (68\% confidence interval). Adapted from \citep{Myhre2013}}\label{fig:O3budget}
\end{figure}

@@

\subsubsection*{Ozone tropospheric chemistry}\label{intro_O3chem}
Keeping in mind the role of tropospheric ozone, a necessary step is to revisit the chemistry that controls its abundance.

The key chemical processes that lead to $O_3$ formation and destruction in the troposphere are driven by reaction cycles involving free-radical intermediates, which are mainly formed from the photolysis of $O_3$ itself (\ref{reac:O3+hv=O1D+O2}) \citep{Fowler2008}. At UV wavelengths shorter than 320 nm, $O_3$ photolysis generates excited oxygen atoms ($O(^1D)$), which than can either react with water vapour to form $OH$ radicals (\ref{reac:O1D+H2O=OH+OH}), or collide with an inert molecule (denoted '$M$') to reform $O_3$ (\ref{reac:O1D=O3P} and \ref{reac:O3P+O2=O3}):
\begin{equation}\label{reac:O3+hv=O1D+O2}
O_3 + hv \rightarrow O(^1D) + O_2
\end{equation}
\begin{equation}\label{reac:O1D+H2O=OH+OH}
O(^1D) + H_2O \rightarrow OH + OH
\end{equation}
\begin{equation}\label{reac:O1D=O3P}
O(^1D) \xrightarrow{M} O(^3P)
\end{equation}
\begin{equation}\label{reac:O3P+O2=O3}
O(^3P) + O_2 \xrightarrow{M} O_3
\end{equation}
Once formed the OH radicals mainly react with $CO$ and $CH_4$ and initiate the $O_3$ formation or removal cycles by producing the peroxy radicals ($HO_2$, $CH_3O_2$):
\begin{equation}\label{reac:OH+CO=H+CO2}
OH + CO \rightarrow H + CO_2
\end{equation}
\begin{equation}\label{reac:H+O2=HO2}
H + O_2 \xrightarrow{M} HO_2
\end{equation}
\begin{equation}\label{reac:OH+CH4=CH3+H2O}
OH + CH_4 \rightarrow CH_3 + H_2O
\end{equation}
\begin{equation}\label{reac:CH3O+O2=CH3O2}
CH_3 + O_2 \xrightarrow{M} CH_3O_2
\end{equation}
The fate of the peroxy radicals depends on the presence of $NO_x$ ($=NO + NO_2$). In low-$NO_x$ conditions (remote regions), the $HO_2$ radical reacts with another $HO_2$ to form hydrogen peroxide ($H_2O_2$) or reacts with the methyl peroxy radical ($CH_3O_2$) to form methyl hydrogen peroxide ($CH_3OOH$):
\begin{equation} \label{reac:HO2+HO2=H2O2+O2}
HO_2 + HO_2 \rightarrow H_2O_2 + O_2
\end{equation}
\begin{equation} \label{reac:HO2+CH3O2=CH3OOH+O2}
HO_2 + CH_3O_2 \rightarrow CH_3OOH + O_2
\end{equation}
On a whole, in low-$NO_x$ conditions, a net $O_3$ destruction takes place, because the reaction sequence is initiated by $O_3$ photolysis. Plus in the same conditions some additional $O_3$ removal can occur because of the reaction of the $HO_2$ radical with $O_3$ leading to further destruction of ozone in a chain sequence involving formation of $OH$ radicals:
\begin{equation}\label{reac:HO2+O3=OH+2O2}
HO_2 + O_3 \rightarrow OH + 2O_2
\end{equation}
\begin{equation}\label{reac:OH+O3=HO2+O2}
OH + O_3 \rightarrow HO_2 + O_2
\end{equation}

In intermediate $NO_x$ conditions (rural areas of most industrialised countries), the peroxy radicals react with nitrogen monoxide ($NO$) to form nitrogen dioxide ($NO_2$), which subsequent photolysis leads to $O_3$ generation:
\begin{equation}\label{reac:HO2+NO=NO2+OH}
HO_2 + NO \rightarrow NO_2 + OH
\end{equation}
\begin{equation}\label{reac:CH3O2+NO=NO2+CH3O}
CH_3O_2 + NO \rightarrow NO_2 + CH_3O
\end{equation}
\begin{equation}\label{reac:NO2+hv=NO+O3P}
NO_2 + hv \rightarrow NO + O(^3P)
\end{equation}
\begin{equation}
O(^3P) + O_2 \xrightarrow{M} O_3\tag{\ref{reac:O3P+O2=O3}}
\end{equation}
As shown in Figure \ref{fig:O3scheme}, the reactions (\ref{reac:HO2+NO=NO2+OH}) and (\ref{reac:CH3O2+NO=NO2+CH3O}) produce parts of the free-radical propagated $O_3$ forming cycles, which may occur a number of times before being halted by radical termination reactions (\ref{reac:HO2+HO2=H2O2+O2} and \ref{reac:HO2+CH3O2=CH3OOH+O2}), and the higher $NO_x$ level is the greater number of these cycles can occur. Therefore, in intermediate $NO_x$ conditions a net $O_3$ formation takes place and it is sensitive to $NO_x$.

\begin{figure}[h!]
\centering
\input{diss_NewO3scheme}
\caption{$O_3$ chemistry scheme}\label{fig:O3scheme}
\end{figure}

In high-$NO_x$ conditions (urban environment), the $OH$ radical may have another fate rather than reacting with $CO$, $CH_4$ and $O_3$. Depending on the concentration of $NO_2$ it can react with it to form nitric acid ($HNO_3$):
\begin{equation}\label{reac:OH+NO2=HNO3}
OH + NO_2 \xrightarrow{M} HNO_3
\end{equation}
The formation of $HNO_3$ terminates free-radical propagated $O_3$ forming cycles leading to a decrease in $O_3$ production rate. However, elevated emissions of $CO$ and $CH_4$ (and other nonmethane volatile organic compounds (nmVOCs)) may make reactions (\ref{reac:OH+CO=H+CO2})-(\ref{reac:OH+CH4=CH3+H2O}) more competitive with the reaction (\ref{reac:OH+NO2=HNO3}) and lead to an increase in the $O_3$ production rate. Therefore, in high-$NO_x$ conditions the $O_3$ production rate can either decrease or increase depending on $NO_x$ and VOCs levels.

The free-radical propagated oxidation of hydrocarbons leads to the generation of carbonyl compounds, among which formaldehyde ($HCHO$) is the most common. Its photolysis and further oxidation contributes to radical generation:
\begin{equation}\label{reac:HCHO+hv=H+HCO}
HCHO + hv \rightarrow H + HCO
\end{equation}
\begin{equation}\label{reac:HCO+O2=HO2+CO}
HCO + O_2 \rightarrow HO_2 + CO
\end{equation}
\begin{equation}
H + O_2 \xrightarrow{M} HO_2\tag{\ref{reac:H+O2=HO2}}
\end{equation}

therefore also having an impact on the rates of $O_3$ production \citep{Fowler2008} as do some other species, which are not mentioned in this section. Their role is described with a required depth in the sections below.

\subsection{Alkyl nitrates}\label{sec:intro_AN}
Alkyl nitrates ($RONO_2$) are organic trace gases, which through one of their formation mechanisms are closely linked with the production of tropospheric ozone \citep{Reeves2007}. This mechanism is presented in the following section.
\subsubsection*{Alkyl nitrates chemistry}\label{into_ANchem}
The photochemical formation of the alkyl nitrates starts when chemistry of the troposphere reaches the regime of $O_3$ production, i.e. when reactions of the peroxy radicals with $NO$ start to compete with their self- and cross-reactions more effectively due to an increase in $NO_x$ mixing ratio. In this case the reaction between the alkyl peroxy radical ($RO_2$, where $R$ denotes methyl group, $CH_3$) and $NO$ can reveal its two-channel nature \citep{Day2003}:
\begin{subequations} \label{reac:RONO2alpha}
\begin{align}
RO_2 + NO &\xrightarrow{1-\alpha_3} RO + NO_2 \label{reac:RO2+NO=aphaRO+NO2}\\
RO_2 + NO &\xrightarrow{\alpha_3} RONO_2 \label{reac:RO2+NO=aphaRONO2}
\end{align}
\end{subequations}
The latter reaction can serve as a termination step of the propagated $O_3$ forming cycle (Figure \ref{fig:O3scheme}), and as \cite{Roberts1998} estimated it is likely to begin to occur when the $NO$ mixing ratio increases up to 100-200 ppt. The yield of alkyl nitrate formation is controlled by the branching ratio ($\alpha_3$), which is temperature and pressure dependent and increases with the number of carbons in the alkyl peroxy radical \citep{Roberts1990}.

An alternative photochemical pathway of the formation of alkyl nitrates is the reaction of the alkoxy radical ($RO$) with $NO_2$:
\begin{equation}\label{reac:RO+NO2=RONO2}
RO + NO_2 \rightarrow RONO_2
\end{equation}
This pathway has been shown to become dominant over reaction (\ref{reac:RO2+NO=aphaRONO2}) only for the production of lower (methyl ($C_1$), ethyl ($C_2$)) nitrates when $NO_x$ mixing ratio exceeds 35 ppb \citep{Archibald2007}, i.e. in highly polluted environments.

@@

\section{Overall objective and specific aims}\label{sec:objams}
Despite of all of the aforesaid, typically chemical mechanisms of global atmospheric models do not contain enough information about some secondary organic compounds \citep{Sommariva2008}, and alkyl nitrates are among them. Underestimation of their concentrations could potentially lead to errors in the estimates of ozone production in remote regions, or to ineffective ozone-control policies. Therefore, the overall objective of this research is to investigate the impact of alkyl nitrate chemistry on tropospheric ozone production by using a chemical mechanism that represents the chemistry related to the problem.

@@

\subsection{Chemical mechanism}\label{sec:method_chem}

@@

\bibliography{diss_refs}
\newpage
\appendix
\section{Model code} \label{sec:appendix1}

\input{model_code}

>>>>

\documentclass[11pt,a4paper]{article}
% Font
\usepackage{helvet}
\renewcommand{\familydefault}{\sfdefault}
% Encoding and language
\usepackage[utf8]{inputenc}
\usepackage[english]{babel}
% Colors
\usepackage[usenames,dvipsnames]{color}
\usepackage{colortbl}
% Paper size and margins
\usepackage{pdflscape}
\usepackage{vmargin}
\setmarginsrb{2.5cm}{2cm}{2.5cm}{2cm}{0cm}{0cm}{0cm}{1.5cm}
\usepackage{pdflscape}
\usepackage{afterpage}
% Indenting and line spacing
\linespread{1.5}
\usepackage{indentfirst}
\setlength{\parskip}{0.5cm}
% Symbols
\usepackage{amsmath}
\usepackage{amsfonts}
\usepackage{amssymb}
% Equations
\usepackage{mathtools} % for arrows
\usepackage{eulervm} % upright font in eqns (Palatino)
\usepackage{cool}
\newcounter{matriz}
\newenvironment{matriz}{\refstepcounter{matriz}\equation}{\tag{\thematriz}\endequation}
\renewcommand{\theequation}{R\arabic{equation}}
% References
\usepackage{natbib}
\bibliographystyle{apalike}
% Hyperlinks and labels
\usepackage[linktocpage=true,plainpages=false,pdfpagelabels=false]{hyperref}
\definecolor{citecolor}{rgb}{0.2,0.7,0.9}
\definecolor{urlcolor}{rgb}{0,0,1}
\hypersetup{
    colorlinks, linkcolor={Blue},
    citecolor={Blue}, urlcolor={Black}
}
% Figures
\usepackage{subcaption}
\usepackage{alphalph}
\usepackage{tikz}
\usetikzlibrary{shapes,arrows}
\usetikzlibrary{positioning}
\usetikzlibrary{calc}
\usepackage{epstopdf}
\usepackage{wrapfig}
% Tables
\usepackage{multirow}
\usepackage{tabularx}
\usepackage{array}
\usepackage{ragged2e}
\newcolumntype{P}[1]{>{\RaggedRight\hspace{0pt}}p{#1}}
% Tikz styles
\tikzstyle{reg} = [rounded rectangle, text width=2cm, fill=black!5, align=center, anchor=west,font=\sffamily\Large\bfseries, minimum height=1cm, inner sep=0]
\tikzstyle{rad} = [rounded rectangle, draw=none, text width=2cm, fill=red!10, align=center, anchor=west,font=\sffamily\Large\bfseries, minimum height=1cm, inner sep=0]
\tikzstyle{long} = [rounded rectangle, draw=none, text width=3.5cm, fill=black!5, align=center, anchor=west,font=\sffamily\Large\bfseries, minimum height=1cm, inner sep=0]   
\tikzstyle{add} = [scale=1.5, draw=none,fill=none,align=center]
\tikzstyle{emp} = [draw=none,fill=none]
\tikzstyle{ar} = [->,draw=black!70,line width=2]
\tikzstyle{ln} = [-,draw=black!70,line width=2]
% Title etc
\newcommand{\mytitle}{Investigation of the relationship between tropospheric ozone production efficiency and carbon bond emissions}
\newcommand{\authorname}{Maria Zamyatina}
\newcommand{\authornumber}{Student No: 100106685}
\newcommand{\degname}{Master of Science}
\newcommand{\univname}{University of East Anglia}
\newcommand{\univaddr}{University Plain \\ Norwich NR47TJ UK}
\newcommand{\depname}{School of Environmental Sciences}
\newcommand{\mydate}{2015}
\newcommand{\figspath}{figures/}
% PDF meta-data
\hypersetup{pdftitle={\mytitle}}
\hypersetup{pdfauthor=\authorname}

\title{\mytitle}

\begin{document}

\input{titlepage}

@@

\newpage
\tableofcontents

\newpage
\section{Introduction} \label{sec:intro}
The Earth’s atmosphere is the recipient of a vast range of chemical compounds emitted by natural and anthropogenic sources. The oxidation of these compounds is the main process taking place in the atmosphere that keeps their concentrations at levels that do not distort the chemical balance of the atmosphere \citep{Prinn2003}. In other words, oxidation is a removal 'cleansing' mechanism for these compounds (various pollutants and greenhouse gases), and its rate is often referred to as the oxidizing capacity of the atmosphere. This capacity depends mainly on two factors: climate and the global burden of the principal oxidants such as ozone ($O_3$), the hydroxyl radical ($OH$) and hydrogen peroxide ($H_2O_2$) \citep{Prinn2003,Thompson1992}.

\subsection{Tropospheric ozone} \label{sec:intro_O3}
Tropospheric ozone ($O_3$) plays an important role in the chemical and thermal balance of the atmosphere. Although it accounts only for about 10\% of the total ozone, it is a primary precursor to the formation of hydroxyl radicals ($OH$) which in association with hydrogen peroxide ($H_2O_2$) determines the 'cleansing' capacity of the atmosphere \citep{Prinn2003,Tarasick2008,Thompson1992}. On the other hand, being a source of main pollutant removal agents, ozone is an air pollutant itself. Long exposure to its high concentrations causes respiratory problems in humans and damage to sensitive plant species \citep{Fowler2008}. Last but not least characteristic of tropospheric ozone is that it is an important greenhouse gas with a current estimated radiative forcing of $0.40\pm 0.20~Wm^{–2}$ which is approximately one fifth of the $CO_2$ radiative forcing \citep{Hartmann2013,Myhre2013}. Due to these reasons a considerable interest in quantifying tropospheric ozone concentrations, budget and trends exists for more than a 100 years \citep{Becker2004}.

Unlike many other air pollutants, tropospheric ozone is not directly emitted, but produced following the oxidation of carbon monoxide ($CO$), methane ($CH_4$), and nonmethane volatile compounds (nmVOCs) in the presence of nitrogen oxides ($NO_x$) \citep{Crutzen1973,Myhre2013}. The emissions of these precursors arise from both natural and anthropogenic sources which have intensified since pre-industrial times \citep{Parrish2014,Volz1988} (Figure \ref{fig:Isaksen2009_fig15}). Presumably it has led to roughly a doubling in background tropospheric ozone concentration \citep{Guicherit2000,Hartmann2013,Tarasick2008,Vingarzan2004}, although contribution from another major source, transport from the stratosphere, may also took place \citep{Fowler2008}. Climate is also believed to be a key driver of ozone production and destruction since the chemical reactions affecting ozone concentration are strongly dependent on such climatic factors as temperature, rainfall and humidity. Therefore in a changing climate tropospheric ozone is not anymore a local air quality issue, but is a global pollution problem \citep{Fowler2008}.

\begin{figure}[h]
\centering
\includegraphics[width=\linewidth]{\figspath Isaksen2009_fig15.png}
\caption{Observed surface ozone at different Northern Hemispheric surface stations \citep{Isaksen2009}.}
\label{fig:Isaksen2009_fig15}
\end{figure}

According to the Intergovernmental Panel on Climate Change (IPCC) Fifth Assessment report (AR5) in the present conditions (around year 2000) the global mean tropospheric ozone budget is approximately 331 Tg \citep{Myhre2013}. However, despite general agreement on how the drivers impact ozone concentrations at global and regional scales, there is considerable inconsistency in its budget between different modelling and observational studies \citep{Stevenson2006,Wild2007,Young2012}. Figure \ref{fig:o3_budget_studies} demonstrates that the range of global tropospheric ozone budget estimates is approximately 60 Tg which is 18\% of the mean. In terms of ozone production the variation is even higher and accounts for 28\% of the corresponding value. In that regard, two questions arise: why does the tropospheric ozone budget differ between models and observations? And secondly, why does it vary to this extent, especially in terms of production? The only way to answer these questions is to perform a rigorous investigation of the factors that drive tropospheric ozone and attribute those differences to respective factors \citep{Young2012}.

\begin{figure}[t]
\centering
\includegraphics[width=0.8\linewidth]{\figspath /o3_budget_studies/o3_budget_studies.pdf}
\caption{Summary of tropospheric ozone global budget model and observation estimates for present (about 2000) conditions. All uncertainties quoted as 1 standard deviation (68\% confidence interval). Adapted from \citep{Myhre2013}.}
\label{fig:o3_budget_studies}
\end{figure}

@@

\subsubsection*{Ozone tropospheric chemistry} \label{intro_O3chem}
Keeping in mind the role of tropospheric ozone, a necessary step is to revisit the chemistry that controls its abundance.

The key chemical processes that lead to $O_3$ formation and destruction in the troposphere are driven by reaction cycles involving free-radical intermediates, which are mainly formed from the photolysis of $O_3$ itself (\ref{reac:O3+hv=O1D+O2}) \citep{Fowler2008}. At UV wavelengths shorter than 320 nm, $O_3$ photolysis generates excited oxygen atoms ($O(^1D)$), which than can either react with water vapour to form $OH$ radicals (\ref{reac:O1D+H2O=OH+OH}), or collide with an inert molecule (denoted '$M$') to reform $O_3$ (\ref{reac:O1D=O3P} and \ref{reac:O3P+O2=O3}):
\begin{equation} \label{reac:O3+hv=O1D+O2}
O_3 + hv \rightarrow O(^1D) + O_2
\end{equation}
\begin{equation} \label{reac:O1D+H2O=OH+OH}
O(^1D) + H_2O \rightarrow OH + OH
\end{equation}
\begin{equation} \label{reac:O1D=O3P}
O(^1D) \xrightarrow{M} O(^3P)
\end{equation}
\begin{equation} \label{reac:O3P+O2=O3}
O(^3P) + O_2 \xrightarrow{M} O_3
\end{equation}
Once formed the OH radicals mainly react with $CO$ and $CH_4$ and initiate the $O_3$ formation or removal cycles by producing the peroxy radicals ($HO_2$, $CH_3O_2$):
\begin{equation} \label{reac:OH+CO=H+CO2}
OH + CO \rightarrow H + CO_2
\end{equation}
\begin{equation} \label{reac:H+O2=HO2}
H + O_2 \xrightarrow{M} HO_2
\end{equation}
\begin{equation} \label{reac:OH+CH4=CH3+H2O}
OH + CH_4 \rightarrow CH_3 + H_2O
\end{equation}
\begin{equation} \label{reac:CH3O+O2=CH3O2}
CH_3 + O_2 \xrightarrow{M} CH_3O_2
\end{equation}
The fate of the peroxy radicals depends on the presence of $NO_x$ ($=NO + NO_2$). In low-$NO_x$ conditions (remote regions), the $HO_2$ radical reacts with another $HO_2$ to form hydrogen peroxide ($H_2O_2$) or reacts with the methyl peroxy radical ($CH_3O_2$) to form methyl hydrogen peroxide ($CH_3OOH$):
\begin{equation} \label{reac:HO2+HO2=H2O2+O2}
HO_2 + HO_2 \rightarrow H_2O_2 + O_2
\end{equation}
\begin{equation} \label{reac:HO2+CH3O2=CH3OOH+O2}
HO_2 + CH_3O_2 \rightarrow CH_3OOH + O_2
\end{equation}
On a whole, in low-$NO_x$ conditions, a net $O_3$ destruction takes place, because the reaction sequence is initiated by $O_3$ photolysis. Plus in the same conditions some additional $O_3$ removal can occur because of the reaction of the $HO_2$ radical with $O_3$ leading to further destruction of ozone in a chain sequence involving formation of $OH$ radicals:
\begin{equation} \label{reac:HO2+O3=OH+2O2}
HO_2 + O_3 \rightarrow OH + 2O_2
\end{equation}
\begin{equation} \label{reac:OH+O3=HO2+O2}
OH + O_3 \rightarrow HO_2 + O_2
\end{equation}

In intermediate $NO_x$ conditions (rural areas of most industrialised countries), the peroxy radicals react with nitrogen monoxide ($NO$) to form nitrogen dioxide ($NO_2$), which subsequent photolysis leads to $O_3$ generation:
\begin{equation} \label{reac:HO2+NO=NO2+OH}
HO_2 + NO \rightarrow NO_2 + OH
\end{equation}
\begin{equation} \label{reac:CH3O2+NO=NO2+CH3O}
CH_3O_2 + NO \rightarrow NO_2 + CH_3O
\end{equation}
\begin{equation} \label{reac:NO2+hv=NO+O3P}
NO_2 + hv \rightarrow NO + O(^3P)
\end{equation}
\begin{equation}
O(^3P) + O_2 \xrightarrow{M} O_3\tag{\ref{reac:O3P+O2=O3}}
\end{equation}
As shown in Figure \ref{fig:o3_chemistry_scheme}, the reactions (\ref{reac:HO2+NO=NO2+OH}) and (\ref{reac:CH3O2+NO=NO2+CH3O}) produce parts of the free-radical propagated $O_3$ forming cycles, which may occur a number of times before being halted by radical termination reactions (\ref{reac:HO2+HO2=H2O2+O2} and \ref{reac:HO2+CH3O2=CH3OOH+O2}), and the higher $NO_x$ level is the greater number of these cycles can occur. Therefore, in intermediate $NO_x$ conditions a net $O_3$ formation takes place and it is sensitive to $NO_x$.

\begin{figure}[h!]
\centering
\includegraphics{\figspath /o3_chemistry_scheme/o3_chemistry_scheme.pdf}
\caption{$O_3$ chemistry scheme.}
\label{fig:o3_chemistry_scheme}
\end{figure}

In high-$NO_x$ conditions (urban environment), the $OH$ radical may have another fate rather than reacting with $CO$, $CH_4$ and $O_3$. Depending on the concentration of $NO_2$ it can react with it to form nitric acid ($HNO_3$):
\begin{equation} \label{reac:OH+NO2=HNO3}
OH + NO_2 \xrightarrow{M} HNO_3
\end{equation}
The formation of $HNO_3$ terminates free-radical propagated $O_3$ forming cycles leading to a decrease in $O_3$ production rate. However, elevated emissions of $CO$ and $CH_4$ (and other nonmethane volatile organic compounds (nmVOCs)) may make reactions (\ref{reac:OH+CO=H+CO2})-(\ref{reac:OH+CH4=CH3+H2O}) more competitive with the reaction (\ref{reac:OH+NO2=HNO3}) and lead to an increase in the $O_3$ production rate. Therefore, in high-$NO_x$ conditions the $O_3$ production rate can either decrease or increase depending on $NO_x$ and VOCs levels.

The free-radical propagated oxidation of hydrocarbons leads to the generation of carbonyl compounds, among which formaldehyde ($HCHO$) is the most common. Its photolysis and further oxidation contributes to radical generation:
\begin{equation} \label{reac:HCHO+hv=H+HCO}
HCHO + hv \rightarrow H + HCO
\end{equation}
\begin{equation} \label{reac:HCO+O2=HO2+CO}
HCO + O_2 \rightarrow HO_2 + CO
\end{equation}
\begin{equation}
H + O_2 \xrightarrow{M} HO_2\tag{\ref{reac:H+O2=HO2}}
\end{equation}
therefore also having an impact on the rates of $O_3$ production \citep{Fowler2008} as do some other species, which are not mentioned in this section. Their role is described with a required depth in the sections below.

\subsection{Alkyl nitrates} \label{sec:intro_AN}
Alkyl nitrates ($RONO_2$) are organic trace gases, which through one of their formation mechanisms are closely linked with the production of tropospheric ozone \citep{Reeves2007}. This mechanism is presented in the following section.

\subsubsection*{Alkyl nitrates chemistry} \label{into_ANchem}
The photochemical formation of the alkyl nitrates starts when chemistry of the troposphere reaches the regime of $O_3$ production, i.e. when reactions of the peroxy radicals with $NO$ start to compete with their self- and cross-reactions more effectively due to an increase in $NO_x$ mixing ratio. In this case the reaction between the alkyl peroxy radical ($RO_2$, where $R$ denotes methyl group, $CH_3$) and $NO$ can reveal its two-channel nature \citep{Day2003}:
\begin{subequations} \label{reac:RONO2alpha}
\begin{align}
RO_2 + NO &\xrightarrow{1-\alpha_3} RO + NO_2 \label{reac:RO2+NO=aphaRO+NO2}\\
RO_2 + NO &\xrightarrow{\alpha_3} RONO_2 \label{reac:RO2+NO=aphaRONO2}
\end{align}
\end{subequations}
The latter reaction can serve as a termination step of the propagated $O_3$ forming cycle (Figure \ref{fig:o3_chemistry_scheme}), and as \cite{Roberts1998} estimated it is likely to begin to occur when the $NO$ mixing ratio increases up to 100-200 ppt. The yield of alkyl nitrate formation is controlled by the branching ratio ($\alpha_3$), which is temperature and pressure dependent and increases with the number of carbons in the alkyl peroxy radical \citep{Roberts1990}.

An alternative photochemical pathway of the formation of alkyl nitrates is the reaction of the alkoxy radical ($RO$) with $NO_2$:
\begin{equation} \label{reac:RO+NO2=RONO2}
RO + NO_2 \rightarrow RONO_2
\end{equation}
This pathway has been shown to become dominant over reaction (\ref{reac:RO2+NO=aphaRONO2}) only for the production of lower (methyl ($C_1$), ethyl ($C_2$)) nitrates when $NO_x$ mixing ratio exceeds 35 ppb \citep{Archibald2007}, i.e. in highly polluted environments.

@@

\section{Overall objective and specific aims} \label{sec:objams}
Despite of all of the aforesaid, typically chemical mechanisms of global atmospheric models do not contain enough information about some secondary organic compounds \citep{Sommariva2008}, and alkyl nitrates are among them. Underestimation of their concentrations could potentially lead to errors in the estimates of ozone production in remote regions, or to ineffective ozone-control policies. Therefore, the overall objective of this research is to investigate the impact of alkyl nitrate chemistry on tropospheric ozone production by using a chemical mechanism that represents the chemistry related to the problem.

@@

\subsection{Chemical mechanism} \label{sec:method_chem}

@@

\bibliography{refs}
\newpage
\appendix
\section{Model code} \label{sec:appendix1}

\input{code}
